---
title: "APIs"
---

<iframe src="https://uio.cloud.panopto.eu/Panopto/Pages/Embed.aspx?id=97266226-eb64-47f8-acec-b02f00de2697&autoplay=false&offerviewer=true&showtitle=true&showbrand=true&captions=false&interactivity=all" height="405" width="720" style="border: 1px solid #464646;" allowfullscreen allow="autoplay"></iframe>

```{r}
#| label: setup
#| eval: true
#| echo: false
#| message: false
#| error: false

library(tidyverse)
person1_url <- "../data/apis/person1.json"
```

Although web pages in `.html` are what we often see with our eyes when using a 
browser, it is not necessarily always the case that this is the best way to 
scrape data. Depending on which website and data you are interested in, there 
are often *back-end* databases from which the websites retrieve information 
based on the user's clicks. Many such websites have an Application Programming 
    Interface (API) available, which you can use relatively freely. And some 
websites are themselves an API. Take for example the 
[Star Wars API](https://swapi.dev/), which is a database of data on characters, 
worlds, movies, etc., in the Star Wars universe.

The front page of SWAPI shows how, for example, you can retrieve data about a 
person:

```{r}
#| label: swapi_intro
#| echo: true
#| eval: false
person1_url <- "https://swapi.dev/api/people/1/"

readLines(person1_url)

```

```{r}
#| label: swapi_intro_hidden
#| echo: false
#| eval: true
#| warning: false

readLines(person1_url)
```


### .json

Here, the data format looks very different than a `.html` because `.html` is a 
bad way to store data. Rather, the vast majority of APIs use formats such as 
`.xml` and `.json`. In SWAPI's case, we get data in `.json` format. This format 
does not lend itself very well to reading with `readLines()`. But, as always, 
someone has made a package that parses data into `.json` for us:

```{r}
#| label: swapi_luke
#| message: false
library(jsonlite)

person1 <- read_json(person1_url)

names(person1)
class(person1)

person1$name
person1$starships

```

Elements such as `starships`, `homeworld` and `films` link to other parts of 
the API, from which further data can be extracted if desired.

Below you will find a slightly longer example of a potential workflow for 
SWAPI, which you can experiment with:

```{r}
#| label: swapi_long_ex
#| file: "../scripts/swapi_ex.R"
#| eval: false
#| code-fold: true
```

:::callout-tip
If you are working with `.json` files that are quite unmanageable, is to use
The `listviewer` package. It provides a very clear visual tree of the data..
:::

### .xml

The second data format most common in APIs is `.xml`. Lets use public transport 
stops in Oslo via the 
[Entur API](https://developer.entur. org/stops-and-timetable-data) API as an 
example. `.xml` is quite similar to `.html`, just easier to work with (mostly).

The first thing we need to do is download data locally to our machine -- there 
is quite a lot of data to work with here. The code snippet below checks if we 
have downloaded the file before and only downloads it if it is not already 
there. We then only need to download the file once -- which is fine in this 
and most cases.

```{r}
#| label: entur_dl
#| eval: false
if(file.exists("../data/apis/ruter.xml") == FALSE){
  download.file(url = "https://api.entur.io/realtime/v1/rest/et?datasetId=RUT",
                destfile = "../data/04-anskaff_tekst/ruter.xml")
}

```

We will use parts of the `.xml` file, which is a bit too large to open in its 
entirety, to find out which stops in Oslo most lines go through. These parts 
look like this:

```{bash}
#| label: entur_preview
#| code-fold: true
#| eval: false
# This is a Unix command that makes .xml files a little nicer when we print 
# them in the console
xmllint --encode utf8 --format ../data/apis/ruter.xml | sed -n 1185,1247p

```

```
<RecordedCalls>
  <RecordedCall>
    <StopPointRef>NSR:Quay:8107</StopPointRef>
    <Order>1</Order>
    <StopPointName>Lillestrøm bussterminal</StopPointName>
    <AimedDepartureTime>2022-08-03T13:50:00+02:00</AimedDepartureTime>
    <ActualDepartureTime>2022-08-03T13:50:00+02:00</ActualDepartureTime>
  </RecordedCall>
  <RecordedCall>
    <StopPointRef>NSR:Quay:9371</StopPointRef>
    <Order>2</Order>
    <StopPointName>Eikeliveien</StopPointName>
    <AimedArrivalTime>2022-08-03T13:52:00+02:00</AimedArrivalTime>
    <ActualArrivalTime>2022-08-03T13:52:00+02:00</ActualArrivalTime>
    <AimedDepartureTime>2022-08-03T13:52:00+02:00</AimedDepartureTime>
    <ActualDepartureTime>2022-08-03T13:52:00+02:00</ActualDepartureTime>
  </RecordedCall>
  . . .
</RecordedCalls>
```

It is somewhat similar to `.html` in writing, but is a lot more structured.

The next thing we need to do is read the local `.xml` file. We do that with the 
same function we use on front-end `.html` pages: `rvest::read_html()`:

```{r}
#| label: entur_les
#| eval: false
library(rvest)

ruter <- read_html("../data/04-anskaff_tekst/ruter.xml")

```


```{r , echo=FALSE, eval=TRUE}
#| label: entur_jukselitt
#| echo: false
#| eval: true
load("../data/apis/entur_alle_stop.rda")
all_stops <- alle_stopp
```

We are now free to extract the data we want from the file. In our case, we want 
all stops on every public transport route in Oslo. These are found within 
`<recordedcall> . . . </recordedcall>`. The code below may seem a little 
advanced at first glance, but a tip to see what happens inside the function can 
be to create the object `x` as the first list element in 
`stop`^[`x <- stop[[1]]`], then execute each line inside the function only on 
this element.

```{r}
#| label: entur_strukt
#| eval: false
#| echo: true
stops <- ruter %>% html_elements("recordedcall")                           # <1>

all_stops <- lapply(stops, function(x){                                   # <2>
  tibble::tibble(
    stop_id = x %>% html_elements("stoppointref") %>% html_text(),         # <3>
    order = x %>% html_elements("order") %>% html_text(),                  # <4>
    stopp_name = x %>% html_elements("stoppointname") %>% html_text(),     # <5>
    aimed_dep = x %>% html_elements("aimeddeparturetime") %>% html_text(), # <6>
    actual_dep = x %>% html_elements("actualdeparturetime") %>% html_text()# <7>
  )

})

alle_stopp <- bind_rows(alle_stopp)                                        # <8>

```

1. Splits the .xml document into each part contained within `<recordedcall> . . . </recordedcall>`
2. For each of these elements we create a tibble() with ...
3. Id for the stop
4. Order in route
5. Stop name
6. Expected time of departure
7. Actual time of departure
8. Ties together all the stops

Now we have a dataset we can use for visualization. For instance, a word cloud:

```{r}
#| label: entur_wordcloud

# Viser data
head(all_stops)

# Making a new data frame where...
stop_name_count <- alle_stopp %>%
  count(stopp_name) %>%             # we count stop names
  arrange(desc(n)) %>%              # sort the data by # routes
  filter(nchar(stopp_name) > 3) %>% # remove stops with short names
  slice_max(n = 30, order_by = n)   # include only the 3 most used stops


library(ggwordcloud)

# Setting up random colors
cols <- sample(colors(),
               size = nrow(stop_name_count),
               replace = TRUE)

# Making plot
stop_name_count %>%
  ggplot(., aes(label = stopp_name,
                size = n,
                color = cols)) +
  geom_text_wordcloud_area() +
  scale_size_area(max_size = 10) +
  theme_void()

```

As expected, Jernbanetorget has the most routes going through the stop.

::: callout-important
Always check if the website you are gathering data from has an API! If it does,
you should use it, not the front-end webpage, for scraping.

There are also some data sources that have made available packages in R for
working with specific APIs. For instance, 
[WDI](https://cran.r-project.org/web/packages/WDI/index.html),
[ESS](https://cran.r-project.org/web/packages/essurvey/), and
[stortingscrape](https://cran.r-project.org/web/packages/stortingscrape).
:::